Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-03-31 response-to-reviews.Rmd

**Unfortunately, we've been unable to find many studies examining vertical gradients at the interspecific level, and have included those of which we are aware.**

I find it interesting to try and link the empirical and theoretical knowledge on vertical gradients in forests with what is done in DGCMs and what should be done to improve their accuracy. However, I think there is a big step between the review (part I and II) and the author’s view on the implications for models (Scaling across space and time). I wonder if it would be possible to link a little bit more the different parts, and maybe without going that much in the technical details (cohort, big leaf, multi layers, ...) which I think are sometimes imprecise or maybe wrong. Would it be possible to explain more simply what are the vertical gradients presented in the section II, that are misrepresented today in DGCMs? Maybe by explaining which of the ‘biophysical variables’ and ‘leaf traits’ that you listed, are considered constant when they should be variable vertically?

**Thank you for the comment. We included a section that explains how vertical gradients are represented in models:**

- **Models partition radiation above and within the forest vertical profile (i.e., direct vs. diffuse light) using radiative transfer models or a system of two coupled ordinary differential equations, referred to as a two-stream approximation [@sellersCanopyReflectancePhotosynthesis1985; @fisherVegetationDemographicsEarth2018]. Using the latter method, single canopy layers are divided into sun and shade fractions (e.g., in the Community Land Model, CLM model), while models with multiple vegetative layers can analytically solve the two-stream approximation for each layer. Thus, even in single-layer models, key physiological parameters like $V_{cmax}$ and $J_{max}$ vary, decreasing with increasing cumulative LAI or lower light conditions [Table 2, e.g., @krinnerDynamicGlobalVegetation2005]. **

### 2022-03-31 response-to-reviews.Rmd

I like your figures, but maybe consider homogenizing a little bit the styles for the plots between figures 2 3 and 4.

**This will be done by the New Phytologist illustrators (because it is a Tansley review). We will work with them to ensure that all the figures look good.**

L205-206 Maybe add some references. How many degrees? In your Figure 3 there is not really a difference between Tleaf and Tair when SWR = 0. Given that the transpiration nearly stops at night (high humidity, low conductance) I would expect Tleaf – Tair to be very close to zero.

**We revised and added references to this sentence in question, which now reads: **
**Leaves are often cooler than the air on clear nights due to radiative coupling with the very cold sky, and under some daytime conditions [cloudy skies, high wind speeds, and high $T_{air}$, @vogelLeavesLowestHighest2009; @rey-sanchez_spatial_2016;  @cavaleriColdbloodedForestsWarming2020].**

@@

Figure 3: TLeaf and Tleaf.

**Done**

l258 under soil wet conditions?

**We have reworded this sentence to read, "In contrast, under conditions conducive to stomatal opening, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory limit evaporative cooling  [Fig. `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999; @leighInfluenceLeafSize2017]."**

### 2022-03-31 vertical-thermal-review.Rmd
small update

Elevation of $T_{leaf}$ above $T_{air}$ is not uniform across time and space, but varies with micrometeorological conditions. During drought (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, $T_{leaf} - T_{air}$ will tend to be elevated in the upper canopy compared to the understory (Fig. `r fig_leaf_T`).
This is because sun leaves are exposed to higher irradiance and VPD, and therefore have stronger $g_s$ limitation than shade leaves [Fig. `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018]. 
In contrast, under conditions conducive to stomatal opening, higher wind speeds and $g_b$ enable cooling in the upper canopy, limit evaporative cooling  [Fig. `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020]. 
Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is less conducive to dissipating excess heat compared to upper-canopy leaves. 
Limitations in heat dissipation in shade leaves, together with physiological propensities, may cause above-optimal $T_{leaf}$ under slight heat or drought stress or during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013].
In addition to lower wind speeds (Fig. `r fig_NEON_vertical`), higher RH in the understory and inner canopy (Fig. `r fig_NEON_vertical`) also inhibits evaporative cooling by reducinf VPD, thus increasing $T_{leaf}$-$T_{air}$ [Fig. `r fig_leaf_T`c, @tibbittsHumidityPlants1979; @songTreeSurfaceTemperature2020;  @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021a; @dietzVerticalPatternsDuration2007].

>>>>

Elevation of $T_{leaf}$ above $T_{air}$ is not uniform across time and space, but varies with micrometeorological conditions. During drought (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, $T_{leaf} - T_{air}$ will tend to be elevated in the upper canopy compared to the understory (Fig. `r fig_leaf_T`).
This is because sun leaves are exposed to higher irradiance and VPD, and therefore have stronger $g_s$ limitation than shade leaves [Fig. `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018]. 
In contrast, under conditions conducive to stomatal opening, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory limit evaporative cooling  [Fig. `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020]. 
Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is less conducive to dissipating excess heat compared to upper-canopy leaves. 
Limitations in heat dissipation in shade leaves, together with physiological propensities, may cause above-optimal $T_{leaf}$ under slight heat or drought stress or during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013].
In addition to lower wind speeds (Fig. `r fig_NEON_vertical`), higher RH in the understory and inner canopy (Fig. `r fig_NEON_vertical`) also inhibits evaporative cooling by reducinf VPD, thus increasing $T_{leaf}$-$T_{air}$ [Fig. `r fig_leaf_T`c, @tibbittsHumidityPlants1979; @songTreeSurfaceTemperature2020;  @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021a; @dietzVerticalPatternsDuration2007].
